The Changing Shape of Empire: A Morphological Study of Curated Chimú Jars and Bottles

### 2022-11-02 ms/ms.tex
<init ms>

---
title: |
  The Changing Shape of Empire: A Morphological Study of Curated Chimú Jars and Bottles 
type: ORIGINAL RESEARCH ARTICLE
author:
  - name: R. Alan Covey
    affil: a
    email: R.Alan.Covey@austin.utexas.edu
  - name: Robert Z. Selden, Jr.
    affil: b
    email: zselden@sfasu.edu
  - name: Astrid Runggaldier
    affil: c
    email: astrid@austin.utexas.edu
  - name: Nicole D. Payntar
    affil: a
    email: npayntar@utexas.edu
affiliation:
  - num: a
    address: |
      Department of Anthropology, The University of Texas at Austin
  - num: b
    address: |
      Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; and Cultural Heritage Department, Jean Monnet University 
  - num: c
    address: |
      Department of Art and Art History, The University of Texas at Austin; The Mesoamerica Center, The University of Texas at Austin
  - num: d
    address: |
      Department of Anthropology, The University of Texas at Austin
bibliography: interactcadsample.bib
appendix: appendix.tex
abstract: |
  After centuries of looting along Peru’s North Coast, archaeologists acknowledge that the pottery of the Chimú Empire is one of the most collected Andean artifacts, but also one of the most poorly understood.  Much of the enduring classificatory uncertainty comes from the problematic provenance of most Chimú vessels, and the fact that the distinctive blackware identified as Chimú represents the production of workshops from across an extensive area, during periods of regional political decentralization (c. 900-1200 CE), imperial growth (c. 1200-1450), and foreign conquest by the Inca (c. 1450-1535) and Spanish (after 1532) empires.  This chapter builds on previous seriations and field observations, using geometric morphometric analysis of a sample of 3D-scanned Chimú bottles from publicly held collections at the University of Texas (Austin) and the American Museum of Natural History.  Since Chimú bottles were formed in workshops using molds, variations in vessel shape can serve as indicators of variable practices.  We compare a sample of common Chimú blackware bottles with a sample of _Inca-Chimú_ vessels that carry features typical of the short Inca occupation of the North Coast.  Differences between the two samples offer new lines of evidence that can guide more precise classifications of these vessels, as well as new interpretations of the imperial history of Peru’s North Coast.
keywords: |
  Sections; lists; figures; tables; mathematics; fonts; references; appendices
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

# The Collecting History of an Imperial Ware

The distinctive blackware pottery of the Chimú Empire of Peru’s north coast is one of the most intensively collected Andean styles, but also one of the least-defined.  Thousands of well-preserved vessels occupy the shelves of museums in South America, North America, and Europe, from the vast collections of famous institutions like Berlin’s Ethnologisches Museum to local museums that possess a single vessel of unknown provenance.  It is impossible to know when Europeans first began to collect Chimú blackwares, although published illustrations date back as far as the early eighteenth century (Figure 1; Frézier 1716).

[Figure 1 about here.  Frézier image of north coast blackwares in an image of “Inca” culture]

By that time, Spaniards had already been plundering Peru’s coastal huacas for nearly two centuries, mining them to extract the gold and silver found in rich burials.  Nineteenth-century collecting shifted the emphasis from precious metals, toward the recovery of well-preserved human crania that could be used to buttress European beliefs in their own racial superiority.  Over time, Chimú ceramics (and other artifacts) were caught up in the new collecting regime as public museums grew in the global North (e.g., Hrdliča 1914:50-51).  By the early twentieth century, large numbers of unauthorized diggers were working full-time excavating tombs at Chan Chan, the ruined Chimú capital (Kroeber 1926:15), and Chimú pottery could be purchased from shops in the Peruvian city of Trujillo (Figure 2).  Beuchat (1918:570) noted at that time that the blackware vessels of Peru’s north coast were already “very well represented in most ethnographic collections.”   Since then, coastal blackwares have continued to be excavated and trafficked by illicit diggers who target more valuable metal artifacts and polychrome textiles (e.g., Atwood 2004), and today they can be readily found at internet auction sites for modest prices.

[Figure 2 about here.  Map of the north coast of Peru]

The history surrounding the collection of Chimú blackwares helps to explain why this ubiquitous artifact class remains poorly understood by professional archaeologists.  When viewed as a byproduct of other collecting activities that unfolded over almost a half millennium, the Chimú ceramic corpus can be seen for what it is: an ambiguous assemblage of mostly orphaned objects.  Almost all Chimú vessels lack a detailed archaeological provenience, and the chain of ownership for most museum pieces can rarely be traced back to Peru.  Because most “Chimú” vessels cannot be affiliated to a particular context in coastal Peru, their identification derives from their stylistic attributes, even though some of the earliest Chimú ceramics documented in archaeological excavations appeared at the Inca site of Machu Picchu (Bingham 1915:269) and the coastal creation shrine of Pachacamac, two sites that lie well beyond the political control of the Chimú Empire (Uhle 1902:756).

This contextual ambiguity intersects with a classificatory problem for Andean archaeology.  Many treat Chimú pottery as typical of the Late Intermediate Period (c. 900-1476) on the north coast of Peru, a huge and undifferentiated period during which the Chimú polity established its capital and local dominion in the Moche Valley, carried out multiple waves of imperial expansion to the north and south, and fell under the hegemony of Inca imperial rule (1476-1532).  Excavators have noted that similar blackwares continued to be produced during the half century of Inca rule, and into the earliest years of the Spanish colonial period.  Without knowing when and where a particular museum object comes from, archaeologists have faced difficulties in subdividing Chimú pottery into phases and substyles, which would enable researchers to study changes and variations within the overall assemblage.

## Early Archaeological Classifications

From early on, classifications of Chimú ceramics focused on vessel attributes.  Beuchat (1918:652-654) noted that coastal tombs had yielded “infinite” formal variations of blackware pots now found in museum collections.  He placed these into five principal categories: (1) globular vessels with rectangular fields of low-relief decoration; (2) geometric vessels (cuboid, ellipsoids, ovular); (3) phytomorphic vessels shaped like squash or fruits; (4) zoomorphic vessels representing a broad range of fauna; and (5) anthropomorphic vessels.  This classification emphasized aesthetic characteristics of vessel shape, rather than the kinds of formal and functional differences seen in later typologies, and Beuchat acknowledged the diversity found within each of his categories, especially the phytomorphic, zoomorphic, and anthropomorphic vessels.  Without any provenience information to guide his analysis, Beuchat did not attempt to seriate Chimú pottery or to distinguish regional variations.

In 1926, Alfred Kroeber made the first concerted effort to describe different phases of Chimú ceramic production, based on a brief visit to the Trujillo area that he made as part of a Field Museum expedition to the central coast.  Kroeber noted previous efforts by Uhle (1902) to differentiate between early and late ceramics from Trujillo, which could be tied to the Moche site of Huaca de la Luna/Huaca del Sol and the Chimú capital of Chan Chan.  Kroeber used the term _Late Chimú_ to refer to Trujillo area blackwares that sometimes were found with Inca ceramics, a designation that encompasses Late Intermediate Period (c. 900-1400s), Inca (1400s-1530s) and some early Colonial pottery.  Based on provenance information that recorded the district where some Late Chimú vessels originated, Kroeber (1926:11) identified a broad distribution that stretched from Piura to Casma, with some vessels found as far south as the Nasca drainage.  To build a typology and sequence of Late Chimú vessels, he studied the pottery from several collections that he purchased in Trujillo on behalf of the Field Museum.

Kroeber’s shape typology for Late Chimú pottery discusses vessels that were common in Moche and Chimú assemblages, and he lists (1926:22) about a dozen Late Chimú shapes, including stirrup-mouth vessels, double jars, double-spout vessels, head-and-spout vessels, figure-and-spout vessels, globular bowls, lipped pots (with and without handles), jars with flaring mouth, flat-handled jars, and jars with tapering spouts.  The list of shapes also includes Inca narrow-mouth jars (_aryballus_) and cups (_kero_) in the Late Chimú assemblage.  In his decorative analysis of vessels in private collections, Kroeber (27-28) identified the _face vase_ and _rotund figure jar_ as additional shapes found in the assemblage.  The discussion of Late Chimú blackware recorded the relative abundance of the most common shapes found in two private collections that he saw while visiting Trujillo.  The Mansiche and Jacoms collections comprised more than 200 vessels, but Kroeber estimated that hundreds of utilitarian pots, plates, and jars had been _rejected_ by the collectors when they purchased vessels from the huaqueros who excavated them near Chan Chan.

In his study of collected ceramics from Trujillo, Kroeber established the outline of a shape typology for Chimú blackware vessels, although it was not a comprehensive representation of ceramics in use on the north coast during the periods of Chimú, Inca, and Spanish imperial expansion.  Building on the earlier observations of the high degree of diversity seen in Chimú blackwares, Kroeber made some important observations that inform subsequent studies.  First, this pottery is widely distributed along the north coast of Peru, although Kroeber (1930:97) argued for strong continuity between regions, which would suggest that regional variations account for little of the overall diversity in the Chimú assemblage.  Second, he noted that blackwares were strongly associated with mortuary contexts, and such pottery was not commonly seen on the surface of the urban core of Chan Chan.  Such vessels were thus more indicative of Chimú funerary practices than everyday life.  Finally, Kroeber described how the private collections that local dealers assembled and sold to museums were already curated in a way that eliminated certain vessel shapes, so that museum collections could be considered representative of the original mortuary assemblage.  As Kroeber noted (1930:95) in a later visit to the north coast, one Trujillo collector “refused to purchase from the huaqueros most of the unhandled jars, exceptions being made in favor of effigy pieces, or occasional plain ones when several vessels were bought in a lot in order to acquire one or two attractive ones.”

## Attempts at Seriation

In the years following Kroeber’s work, analysis of Chimú ceramics continued to be oriented around the integration of regional Andean sequences, and the pottery of the Chimú Empire continued to be discussed alongside Moche ceramics that were considered to be Early Chimú (e.g., Bennett 1937; Willey 1945).  It was not until the 1960s that a concerted effort was made to move beyond the master sequence, to developed a more fine-grained description of the assemblage associated with the Chimú Empire as it developed, expanded, and fell under Inca and Spanish dominion.  In 1966, Harry Scheele and Thomas Patterson published a “preliminary seriation” of Chimú ceramics.  They noted (1966:15) that one reason that so little work had been done to classify such material was its abundance: “probably no other pottery style from the Americas is so well represented on the shelves of museums and private collectors throughout the world.”

@@

Building on preliminary work by Pedro Rojas Ponce and Dorothy Menzel, Sheele and Patterson (1966) compared illustrations from the published literature with vessels from Peru’s Museo Nacional de Antropología y Arqueología and Harvard University’s Peabody Museum of Archaeology and Ethnology.  They created a preliminary seriation of Chimú ceramics from the end of the Middle Horizon to the early Colonial Period, distinguishing seven different phases based on the identification of different vessel shapes, decorative elements, and firing characteristics.  Although the seriation was developed to build a more fine-grained relative chronology, the small sample sizes of some phases presented interpretive challenges.  For example, the fourth phase _Lambayeque_, was based on seven vessels from a single grave in the Lambayeque Valley, located approximately 175 km to the north of Chan Chan, and probably reflects regional ceramic diversity more than a distinct period of Chimú production in the Moche Valley.  Other phases were based on small sample sizes and vessels with limited provenance information.  One important distinction in the Scheele and Patterson seriation was between Chimú ceramics of the late LIP (Chimu Phase T-1) and those from the time of the Inca occupation (Chimu Phase T-2).  The former was defined based on the prevalence of stirrup-spout vessels and the _less carefully made_ production of mold-produced vessels that seemed to represent a more restricted set of shapes than the preceding phase (Trujillo Phase T-2).  The Chimú-Inca phase maintained elements of existing ceramic production, adding features from Inca vessel shapes.  Scheele and Patterson note (1966:24) that this phase seems to be distributed beyond earlier Chimu political boundaries and economic networks—it has been identified on the south and central coast and in the Cuzco region.  Overall, Scheele and Patterson’s work clarifies how some vessel shapes developed over time, making it possible to assign provenienced gravelots to more specific late prehispanic periods.  It also identified some important aspects of production practices and hybridity that have been important for subsequent archaeological fieldwork.

## Field Archaeology and the Reconstruction of Production Practices

@@

At the end of the 1960s, Moseley and Mackey’s Chan Chan-Moche Valley Project (1969-1975) began to increase the intensity of professional archaeological research, with regional survey work in the lower Moche Valley and mapping and excavations at Chan Chan and other nearby sites.  While not focused on cemetery sites, this project and related studies encountered additional mortuary contexts with well-preserved ceramics (see Donnan and Mackey 1978).  It was not until the 1990s, however, that archaeologists began to identify loci of ceramic production based on the presence of overfired sherds, molds and matrices, stamps, tools, kilns, and other evidence (Donnan 1997; Hayashida 1999; Mackey 2003; Mackey and Sapp 2021; Tschauner 2006, 2009).  Some of the workshops reflect ceramic production practices during the preceding era of Chimú imperial expansion, whereas others were in use during the Inca occupation of the north coast.  Surface collections and excavations in production areas generated multiple lines of material evidence, including large numbers of potsherds that could not easily be assigned to the diverse array of shapes identified in collection-based classifications.

The study of Chimú and Chimú-Inca pottery workshops has focused on two central themes:  specialization and cultural (dis)continuity under foreign rule.  These issues were nascent in many of the earlier collection-based studies, but the careful excavation and analysis of production contexts added invaluable evidence to the discussion.  Molds and matrices encountered at workshops indicate that multiple vessel parts were mass-produced and had to be assembled following a specific sequence, which Tschauner (2006:179) interprets as evidence of specialized production during Chimú times.  Tschauner’s excavations at the Pampa de Burros workshop in the Lambayeque Valley identified heavy use of vertical half-molds to produce jars, canteens or flasks, bottles, and ollas—virtually all of the shapes that were used by Chimú-era settlements in the surrounding valley (2006:182).  He sees the local potters as relatively unskilled producers who relied on more adept makers of ceramic molds, and whose _market-oriented_ work could flexibly adapt to consumer preferences (2006:183-185).  Levine (2011) builds on this idea, noting that some of the variability seen in the Chimú assemblage could come from ceramic producers _mixing and matching_ molds for different vessel parts and adornments to produce new combinations.

For the Inca period, scholars acknowledge a degree of continuity in local potting practices, but also some significant changes.  At Tambo Real and La Viña in the Leche Valley, Hayashida (1999) notes the continuity of local practices for shaping vessels (press molds, paddle and anvil) and firing.  While local jar shapes persisted in Inca times, she observes the presence of a mold for an Inca flared jar neck (1999:345), which could be added to jars that were shaped using local press molds.  Overall, Hayashida sees little evidence of Inca _retraining_ of Chimú potters.

Other Inca-Chimú workshops in the Jequetepeque Valley provide evidence that practices varied at different production locations.  Donnan’s (1997:32-35) surface collections at Cañoncillo encountered dozens of mold fragments, most of which were for shaping the bottom and top hemispheres of ollas.  Donnan identified molds for shaping jars along a vertical axis, and for producing bottles, including the distinctive Inca “aryballoid” shape.  Although he reports that no molds appeared for shaping handles or rims, there were molds for producing stirrup spouts, as well as for humans, birds, and other animals.  The production evidence from this site suggests that while some vessels were largely shaped using molds, some elements were hand-formed and showed variations suggesting the work of multiple potters.  At the Inca administrative center of Farfán, Mackey (2004:336-338) excavated a small workshop that produced large jars (tinajas) for brewing or storage, using the paddle-and-anvil technique.  Finally, the excavation of three workshop patios at El Algarrobal de Moro, a lower-order Chimú administrative center, encountered 25 mold fragments, including molds for a stirrup spout vessel and a miniature jar (Mackey and Sapp 2021:21-22).

## Technological Approaches to Chimú Ceramics

@@

## 3D Scanning and Geometric Morphometrics Analysis of Chimú Pottery

>>>>

% interactcadsample.tex
% v1.03 - April 2017

\documentclass[]{interact}

\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
\usepackage{subfigure}% Support for small, `sub' figures and tables
%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required

\usepackage{natbib}% Citation support using natbib.sty
\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty

\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}{Remark}
\newtheorem{notation}{Notation}

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{hyperref}
\usepackage[utf8]{inputenc}
\def\tightlist{}

\begin{document}

\articletype{ORIGINAL RESEARCH ARTICLE}

\title{The Changing Shape of Empire: A Morphological Study of Curated
Chimú Jars and Bottles}

\author{\name{R. Alan Covey$^{a}$, Robert Z. Selden, Jr.$^{b}$, Astrid
Runggaldier$^{c}$, Nicole D. Payntar$^{a}$}
\affil{$^{a}$Department of Anthropology, The University of Texas at
Austin; $^{b}$Heritage Research Center, Stephen F. Austin State
University; Department of Biology, Stephen F. Austin State University;
and Cultural Heritage Department, Jean Monnet
University; $^{c}$Department of Art and Art History, The University of
Texas at Austin; The Mesoamerica Center, The University of Texas at
Austin; $^{d}$Department of Anthropology, The University of Texas at
Austin}
}

\thanks{CONTACT R. Alan
Covey. Email: \href{mailto:R.Alan.Covey@austin.utexas.edu}{\nolinkurl{R.Alan.Covey@austin.utexas.edu}}, Robert
Z. Selden,
Jr.. Email: \href{mailto:zselden@sfasu.edu}{\nolinkurl{zselden@sfasu.edu}}, Astrid
Runggaldier. Email: \href{mailto:astrid@austin.utexas.edu}{\nolinkurl{astrid@austin.utexas.edu}}, Nicole
D.
Payntar. Email: \href{mailto:npayntar@utexas.edu}{\nolinkurl{npayntar@utexas.edu}}}

\maketitle

\begin{abstract}
After centuries of looting along Peru's North Coast, archaeologists
acknowledge that the pottery of the Chimú Empire is one of the most
collected Andean artifacts, but also one of the most poorly understood.
Much of the enduring classificatory uncertainty comes from the
problematic provenance of most Chimú vessels, and the fact that the
distinctive blackware identified as Chimú represents the production of
workshops from across an extensive area, during periods of regional
political decentralization (c.~900-1200 CE), imperial growth
(c.~1200-1450), and foreign conquest by the Inca (c.~1450-1535) and
Spanish (after 1532) empires. This chapter builds on previous seriations
and field observations, using geometric morphometric analysis of a
sample of 3D-scanned Chimú bottles from publicly held collections at the
University of Texas (Austin) and the American Museum of Natural History.
Since Chimú bottles were formed in workshops using molds, variations in
vessel shape can serve as indicators of variable practices. We compare a
sample of common Chimú blackware bottles with a sample of
\emph{Inca-Chimú} vessels that carry features typical of the short Inca
occupation of the North Coast. Differences between the two samples offer
new lines of evidence that can guide more precise classifications of
these vessels, as well as new interpretations of the imperial history of
Peru's North Coast.
\end{abstract}

\begin{keywords}
Sections; lists; figures; tables; mathematics; fonts; references;
appendices
\end{keywords}

\hypertarget{the-collecting-history-of-an-imperial-ware}{%
\section{The Collecting History of an Imperial
Ware}\label{the-collecting-history-of-an-imperial-ware}}

The distinctive blackware pottery of the Chimú Empire of Peru's north
coast is one of the most intensively collected Andean styles, but also
one of the least-defined. Thousands of well-preserved vessels occupy the
shelves of museums in South America, North America, and Europe, from the
vast collections of famous institutions like Berlin's Ethnologisches
Museum to local museums that possess a single vessel of unknown
provenance. It is impossible to know when Europeans first began to
collect Chimú blackwares, although published illustrations date back as
far as the early eighteenth century (Figure 1; Frézier 1716).

{[}Figure 1 about here. Frézier image of north coast blackwares in an
image of ``Inca'' culture{]}

By that time, Spaniards had already been plundering Peru's coastal
huacas for nearly two centuries, mining them to extract the gold and
silver found in rich burials. Nineteenth-century collecting shifted the
emphasis from precious metals, toward the recovery of well-preserved
human crania that could be used to buttress European beliefs in their
own racial superiority. Over time, Chimú ceramics (and other artifacts)
were caught up in the new collecting regime as public museums grew in
the global North (e.g., Hrdliča 1914:50-51). By the early twentieth
century, large numbers of unauthorized diggers were working full-time
excavating tombs at Chan Chan, the ruined Chimú capital (Kroeber
1926:15), and Chimú pottery could be purchased from shops in the
Peruvian city of Trujillo (Figure 2). Beuchat (1918:570) noted at that
time that the blackware vessels of Peru's north coast were already
``very well represented in most ethnographic collections.'' Since then,
coastal blackwares have continued to be excavated and trafficked by
illicit diggers who target more valuable metal artifacts and polychrome
textiles (e.g., Atwood 2004), and today they can be readily found at
internet auction sites for modest prices.

{[}Figure 2 about here. Map of the north coast of Peru{]}

The history surrounding the collection of Chimú blackwares helps to
explain why this ubiquitous artifact class remains poorly understood by
professional archaeologists. When viewed as a byproduct of other
collecting activities that unfolded over almost a half millennium, the
Chimú ceramic corpus can be seen for what it is: an ambiguous assemblage
of mostly orphaned objects. Almost all Chimú vessels lack a detailed
archaeological provenience, and the chain of ownership for most museum
pieces can rarely be traced back to Peru. Because most ``Chimú'' vessels
cannot be affiliated to a particular context in coastal Peru, their
identification derives from their stylistic attributes, even though some
of the earliest Chimú ceramics documented in archaeological excavations
appeared at the Inca site of Machu Picchu (Bingham 1915:269) and the
coastal creation shrine of Pachacamac, two sites that lie well beyond
the political control of the Chimú Empire (Uhle 1902:756).

This contextual ambiguity intersects with a classificatory problem for
Andean archaeology. Many treat Chimú pottery as typical of the Late
Intermediate Period (c.~900-1476) on the north coast of Peru, a huge and
undifferentiated period during which the Chimú polity established its
capital and local dominion in the Moche Valley, carried out multiple
waves of imperial expansion to the north and south, and fell under the
hegemony of Inca imperial rule (1476-1532). Excavators have noted that
similar blackwares continued to be produced during the half century of
Inca rule, and into the earliest years of the Spanish colonial period.
Without knowing when and where a particular museum object comes from,
archaeologists have faced difficulties in subdividing Chimú pottery into
phases and substyles, which would enable researchers to study changes
and variations within the overall assemblage.

\hypertarget{early-archaeological-classifications}{%
\subsection{Early Archaeological
Classifications}\label{early-archaeological-classifications}}

From early on, classifications of Chimú ceramics focused on vessel
attributes. Beuchat (1918:652-654) noted that coastal tombs had yielded
``infinite'' formal variations of blackware pots now found in museum
collections. He placed these into five principal categories: (1)
globular vessels with rectangular fields of low-relief decoration; (2)
geometric vessels (cuboid, ellipsoids, ovular); (3) phytomorphic vessels
shaped like squash or fruits; (4) zoomorphic vessels representing a
broad range of fauna; and (5) anthropomorphic vessels. This
classification emphasized aesthetic characteristics of vessel shape,
rather than the kinds of formal and functional differences seen in later
typologies, and Beuchat acknowledged the diversity found within each of
his categories, especially the phytomorphic, zoomorphic, and
anthropomorphic vessels. Without any provenience information to guide
his analysis, Beuchat did not attempt to seriate Chimú pottery or to
distinguish regional variations.

In 1926, Alfred Kroeber made the first concerted effort to describe
different phases of Chimú ceramic production, based on a brief visit to
the Trujillo area that he made as part of a Field Museum expedition to
the central coast. Kroeber noted previous efforts by Uhle (1902) to
differentiate between early and late ceramics from Trujillo, which could
be tied to the Moche site of Huaca de la Luna/Huaca del Sol and the
Chimú capital of Chan Chan. Kroeber used the term \emph{Late Chimú} to
refer to Trujillo area blackwares that sometimes were found with Inca
ceramics, a designation that encompasses Late Intermediate Period
(c.~900-1400s), Inca (1400s-1530s) and some early Colonial pottery.
Based on provenance information that recorded the district where some
Late Chimú vessels originated, Kroeber (1926:11) identified a broad
distribution that stretched from Piura to Casma, with some vessels found
as far south as the Nasca drainage. To build a typology and sequence of
Late Chimú vessels, he studied the pottery from several collections that
he purchased in Trujillo on behalf of the Field Museum.

Kroeber's shape typology for Late Chimú pottery discusses vessels that
were common in Moche and Chimú assemblages, and he lists (1926:22) about
a dozen Late Chimú shapes, including stirrup-mouth vessels, double jars,
double-spout vessels, head-and-spout vessels, figure-and-spout vessels,
globular bowls, lipped pots (with and without handles), jars with
flaring mouth, flat-handled jars, and jars with tapering spouts. The
list of shapes also includes Inca narrow-mouth jars (\emph{aryballus})
and cups (\emph{kero}) in the Late Chimú assemblage. In his decorative
analysis of vessels in private collections, Kroeber (27-28) identified
the \emph{face vase} and \emph{rotund figure jar} as additional shapes
found in the assemblage. The discussion of Late Chimú blackware recorded
the relative abundance of the most common shapes found in two private
collections that he saw while visiting Trujillo. The Mansiche and Jacoms
collections comprised more than 200 vessels, but Kroeber estimated that
hundreds of utilitarian pots, plates, and jars had been \emph{rejected}
by the collectors when they purchased vessels from the huaqueros who
excavated them near Chan Chan.

In his study of collected ceramics from Trujillo, Kroeber established
the outline of a shape typology for Chimú blackware vessels, although it
was not a comprehensive representation of ceramics in use on the north
coast during the periods of Chimú, Inca, and Spanish imperial expansion.
Building on the earlier observations of the high degree of diversity
seen in Chimú blackwares, Kroeber made some important observations that
inform subsequent studies. First, this pottery is widely distributed
along the north coast of Peru, although Kroeber (1930:97) argued for
strong continuity between regions, which would suggest that regional
variations account for little of the overall diversity in the Chimú
assemblage. Second, he noted that blackwares were strongly associated
with mortuary contexts, and such pottery was not commonly seen on the
surface of the urban core of Chan Chan. Such vessels were thus more
indicative of Chimú funerary practices than everyday life. Finally,
Kroeber described how the private collections that local dealers
assembled and sold to museums were already curated in a way that
eliminated certain vessel shapes, so that museum collections could be
considered representative of the original mortuary assemblage. As
Kroeber noted (1930:95) in a later visit to the north coast, one
Trujillo collector ``refused to purchase from the huaqueros most of the
unhandled jars, exceptions being made in favor of effigy pieces, or
occasional plain ones when several vessels were bought in a lot in order
to acquire one or two attractive ones.''

\hypertarget{attempts-at-seriation}{%
\subsection{Attempts at Seriation}\label{attempts-at-seriation}}

In the years following Kroeber's work, analysis of Chimú ceramics
continued to be oriented around the integration of regional Andean
sequences, and the pottery of the Chimú Empire continued to be discussed
alongside Moche ceramics that were considered to be Early Chimú (e.g.,
Bennett 1937; Willey 1945). It was not until the 1960s that a concerted
effort was made to move beyond the master sequence, to developed a more
fine-grained description of the assemblage associated with the Chimú
Empire as it developed, expanded, and fell under Inca and Spanish
dominion. In 1966, Harry Scheele and Thomas Patterson published a
``preliminary seriation'' of Chimú ceramics. They noted (1966:15) that
one reason that so little work had been done to classify such material
was its abundance: ``probably no other pottery style from the Americas
is so well represented on the shelves of museums and private collectors
throughout the world.''

@@

Building on preliminary work by Pedro Rojas Ponce and Dorothy Menzel,
Sheele and Patterson (1966) compared illustrations from the published
literature with vessels from Peru's Museo Nacional de Antropología y
Arqueología and Harvard University's Peabody Museum of Archaeology and
Ethnology. They created a preliminary seriation of Chimú ceramics from
the end of the Middle Horizon to the early Colonial Period,
distinguishing seven different phases based on the identification of
different vessel shapes, decorative elements, and firing
characteristics. Although the seriation was developed to build a more
fine-grained relative chronology, the small sample sizes of some phases
presented interpretive challenges. For example, the fourth phase
\emph{Lambayeque}, was based on seven vessels from a single grave in the
Lambayeque Valley, located approximately 175 km to the north of Chan
Chan, and probably reflects regional ceramic diversity more than a
distinct period of Chimú production in the Moche Valley. Other phases
were based on small sample sizes and vessels with limited provenance
information. One important distinction in the Scheele and Patterson
seriation was between Chimú ceramics of the late LIP (Chimu Phase T-1)
and those from the time of the Inca occupation (Chimu Phase T-2). The
former was defined based on the prevalence of stirrup-spout vessels and
the \emph{less carefully made} production of mold-produced vessels that
seemed to represent a more restricted set of shapes than the preceding
phase (Trujillo Phase T-2). The Chimú-Inca phase maintained elements of
existing ceramic production, adding features from Inca vessel shapes.
Scheele and Patterson note (1966:24) that this phase seems to be
distributed beyond earlier Chimu political boundaries and economic
networks---it has been identified on the south and central coast and in
the Cuzco region. Overall, Scheele and Patterson's work clarifies how
some vessel shapes developed over time, making it possible to assign
provenienced gravelots to more specific late prehispanic periods. It
also identified some important aspects of production practices and
hybridity that have been important for subsequent archaeological
fieldwork.

\hypertarget{field-archaeology-and-the-reconstruction-of-production-practices}{%
\subsection{Field Archaeology and the Reconstruction of Production
Practices}\label{field-archaeology-and-the-reconstruction-of-production-practices}}

@@

At the end of the 1960s, Moseley and Mackey's Chan Chan-Moche Valley
Project (1969-1975) began to increase the intensity of professional
archaeological research, with regional survey work in the lower Moche
Valley and mapping and excavations at Chan Chan and other nearby sites.
While not focused on cemetery sites, this project and related studies
encountered additional mortuary contexts with well-preserved ceramics
(see Donnan and Mackey 1978). It was not until the 1990s, however, that
archaeologists began to identify loci of ceramic production based on the
presence of overfired sherds, molds and matrices, stamps, tools, kilns,
and other evidence (Donnan 1997; Hayashida 1999; Mackey 2003; Mackey and
Sapp 2021; Tschauner 2006, 2009). Some of the workshops reflect ceramic
production practices during the preceding era of Chimú imperial
expansion, whereas others were in use during the Inca occupation of the
north coast. Surface collections and excavations in production areas
generated multiple lines of material evidence, including large numbers
of potsherds that could not easily be assigned to the diverse array of
shapes identified in collection-based classifications.

The study of Chimú and Chimú-Inca pottery workshops has focused on two
central themes: specialization and cultural (dis)continuity under
foreign rule. These issues were nascent in many of the earlier
collection-based studies, but the careful excavation and analysis of
production contexts added invaluable evidence to the discussion. Molds
and matrices encountered at workshops indicate that multiple vessel
parts were mass-produced and had to be assembled following a specific
sequence, which Tschauner (2006:179) interprets as evidence of
specialized production during Chimú times. Tschauner's excavations at
the Pampa de Burros workshop in the Lambayeque Valley identified heavy
use of vertical half-molds to produce jars, canteens or flasks, bottles,
and ollas---virtually all of the shapes that were used by Chimú-era
settlements in the surrounding valley (2006:182). He sees the local
potters as relatively unskilled producers who relied on more adept
makers of ceramic molds, and whose \emph{market-oriented} work could
flexibly adapt to consumer preferences (2006:183-185). Levine (2011)
builds on this idea, noting that some of the variability seen in the
Chimú assemblage could come from ceramic producers \emph{mixing and
matching} molds for different vessel parts and adornments to produce new
combinations.

For the Inca period, scholars acknowledge a degree of continuity in
local potting practices, but also some significant changes. At Tambo
Real and La Viña in the Leche Valley, Hayashida (1999) notes the
continuity of local practices for shaping vessels (press molds, paddle
and anvil) and firing. While local jar shapes persisted in Inca times,
she observes the presence of a mold for an Inca flared jar neck
(1999:345), which could be added to jars that were shaped using local
press molds. Overall, Hayashida sees little evidence of Inca
\emph{retraining} of Chimú potters.

Other Inca-Chimú workshops in the Jequetepeque Valley provide evidence
that practices varied at different production locations. Donnan's
(1997:32-35) surface collections at Cañoncillo encountered dozens of
mold fragments, most of which were for shaping the bottom and top
hemispheres of ollas. Donnan identified molds for shaping jars along a
vertical axis, and for producing bottles, including the distinctive Inca
``aryballoid'' shape. Although he reports that no molds appeared for
shaping handles or rims, there were molds for producing stirrup spouts,
as well as for humans, birds, and other animals. The production evidence
from this site suggests that while some vessels were largely shaped
using molds, some elements were hand-formed and showed variations
suggesting the work of multiple potters. At the Inca administrative
center of Farfán, Mackey (2004:336-338) excavated a small workshop that
produced large jars (tinajas) for brewing or storage, using the
paddle-and-anvil technique. Finally, the excavation of three workshop
patios at El Algarrobal de Moro, a lower-order Chimú administrative
center, encountered 25 mold fragments, including molds for a stirrup
spout vessel and a miniature jar (Mackey and Sapp 2021:21-22).

\hypertarget{technological-approaches-to-chimuxfa-ceramics}{%
\subsection{Technological Approaches to Chimú
Ceramics}\label{technological-approaches-to-chimuxfa-ceramics}}

@@

\hypertarget{d-scanning-and-geometric-morphometrics-analysis-of-chimuxfa-pottery}{%
\subsection{3D Scanning and Geometric Morphometrics Analysis of Chimú
Pottery}\label{d-scanning-and-geometric-morphometrics-analysis-of-chimuxfa-pottery}}

After a century of classifying Chimú ceramics and reconstructing their
production and social uses, archaeologists have yet to bring order to
the \emph{infinite} diversity of the museum assemblage, but they have
developed important questions for empirical analysis. Rather than
attempting to fit Chimú vessels into a neat taxonomy, researchers have
drawn attention to the different factors influencing diversity in late
prehispanic blackwares. Blackware ceramics were produced over several
centuries and across a broad stretch of the north coast, and a
significant degree of variation comes from the practices of individual
potters at different workshops, who were producing different vessels to
meet the changing aesthetic tastes of coastal consumers. There appear to
have been differences in the organization of pottery workshops, and some
producers relied more on the use of molds and matrices than others. It
is not presently clear who designed and produced molds and matrices, or
how broadly used and consistently replicated such production templates
were. A fine-grained three-dimensional study of vessel bodies and key
elements known to have been molded in some instances (e.g., stirrup
spouts) could quantify the amount of variation seen in well-preserved
vessels, contributing to the discussion of how hierarchical and
centralized Chimú ceramic production was beyond the local workshop.
